HEP Software Foundation Community White Paper Working Group -- Detector Simulation

### 2018-01-23 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
incorporating Daniel's changes

\abstract{ A working group on detector simulation was formed as part of the HEP 
Software Foundation's initiative to prepare a Community White Paper that describes 
the main software challenges and opportunities to be faced in our field over the next 
decade. The working group met over a period of several months in order to review the 
current status of the Full and Fast simulation applications of HEP experiments 
and the improvements that will need to be made in order to meet the goals of future HEP 
experimental programmes.  The scope of the topics covered includes the main components 
of a HEP simulation application, including MC truth handling, geometry modeling, particle 
propagation in materials and fields, physics modeling of the interactions of particles 
with matter, the treatment of pileup and other backgrounds, as well as signal processing 
and digitisation. The work programme described in this document focuses on the need to 
improve both software and physics performance, in view of the need to improve statistical precision
and to address the diversification of physics programmes.}

@@

For all its success so far, the challenges faced by the HEP field in the
simulation domain are daunting. During the first two runs, the LHC
experiments produced, reconstructed, stored, transferred, and analyzed
tens of billions of simulated events. This effort required more than
half of the total computing resources allocated to the experiments. As
part of the high-luminosity LHC physics program (HL-LHC) and through the
end of the 2030's, the upgraded experiments expect to collect 150 times
more data than in Run 1. The 50 PB of raw data produced in 2016 will
grow to approximately 600 PB in 2026 while the CPU needs will increase
by a factor of about 60. Demand for larger simulation samples to satisfy
analysis needs will grow accordingly. In addition, simulation tools have
to serve diverse communities, including accelerator-based particle
physics research utilizing proton-proton colliders, neutrino and muon
experiments, as well as the cosmic frontier. The complex detectors of
the future, with different module- or cell-level shapes, finer
segmentation, and novel materials and detection techniques, require
additional features in geometry tools and bring new demands on physics
coverage and accuracy within the constraints of the available computing
budget. More extensive use of Fast Simulation is a potential solution,
under the assumption that it is possible to improve time performance
without an unacceptable loss of physics accuracy.

@@

Muon experiments such as Muon g-2 and Mu2e also face large simulation
challenges. Because they are searching for extremely rare effects, they
must grapple with very low signal to background ratios, and the modeling
of low cross-section background processes. Often the physics of these
processes is not well understood, introducing large systematic
uncertainties. Additionally, the size of the computational problem is a
serious challenge, as large simulation runs are required to adequately
sample all relevant areas of experimental phase space, even when
techniques to minimise the required computations are used. Another
aspect related to the background events is the need to simulate the
effects of low energy neutrons, which requires large computational
resources. For example, Mu2e used a total of approximately 60 million
CPU hours, or 83 thousand core months, in a 2016 simulation campaign to
produce the Full Simulation sample needed to complete a technical design
report delivered in 2016. Geant4 is the primary simulation toolkit for
all stages of these experiments. For comparison, the CMS experiment
spent 860 thousand core months on simulation in the May 2015-May 2016
period.

Although the computing needs of the intensity frontier are currently
small compared to the LHC experiments, they are expected to grow
substantially in the next decade, and it is imperative that the
intensity frontier experiments make use of modern computing
architectures and algorithms. This transition will be complicated by the
fact that the intensity frontier is fractured into many small
experiments, some of which have as few as 50 collaborators, limiting the
manpower available for updating detector simulations to use new tools.
Efforts such as LArSoft{[}7{]} have recognised this challenge and aim to
pool resources of many experiments for the benefit of all.

@@

The Geant4 Collaboration has invested significant effort into improving
the toolkit computing performance in the past decade, as code was
reviewed and optimised. In 2013, the release of event-level
multithreading capabilities in Geant4 brought significant memory
savings, as illustrated in Figure 10, which shows the CPU time (top) and
memory (bottom) consumption of a simplified CMS standalone simulation
application, based on Geant4 version 10.3.p01, runs for 50 GeV pions.
Geant4 achieves almost perfect strong scaling (top plot) with the number
of available cores (57) and still gains about +30\% in hyperthreading
regime {[}29,30{]}. The large memory reduction obtained with
multi-threading is clearly visible (bottom). While a multi-process
approach can reach the O(10GB) memory consumption with a relatively
small number of threads (Geant4 version 9.6), it is only with a well
designed multi-threading approach, which shares the read-only data among
threads, that it is possible to run with the maximum number of threads
remaining in a memory budget of approximately 1 GB. Recent versions of
Geant4 have put emphasis on further reducing memory consumption to the
current level of about 10MB needed for each additional thread.

@@

Tables 4 and 5 shows the relative CPU performance of a number of
configurations of the ATLAS simulation application. The largest gain
comes from introducing shower libraries in the forward calorimeter,
which comes at a cost of 430MB of VMEM per job. No other option applied
here has an appreciable effect on memory requirements of the
application. The neutron time cut shows only a small CPU improvement,
but was shown to reduce the size of the output file by almost a factor
of two in previous tests. This file size issue can also be mitigated by
various approaches to hit storage, however. Removing range cuts
decreases the simulation time per event, because the thin volumes of the
sampling calorimeter demand a tighter range cut, including 30um in the
forward calorimeter and 300um in the EM calorimeters, to provide an
accurate physics description. This penalty is not present in the CMS
simulation, since the CMS EM calorimeter has much larger active volumes.
Still, the effect of range cuts on the two applications is similar.

@@

There are many techniques to make simulation faster. Some of them are
applied to Geant4-based Full Simulation applications, such as for
example Russian Roulette and shower libraries, to speedup simulation in
a given sub-detector. In the literature, these applications are still
labeled Full Simulation and Geant4-based. Some experiments have
developed alternative fast simulation frameworks, based on fast track
propagation through a simplified geometry and parameterised showers,
which are less accurate and are typically used in detector upgrade
studies and to produce ``signal'' samples that require scans of a large
region of theory parameter space. Generally, the faster the simulation,
the more imprecise the physics is. Examples of fast simulation
techniques, used to speedup Geant4-based applications or in fast
simulation frameworks, are:

@@

While the focus in this section and in the literature so far has been on
using machine learning to do fast simulation (e.g. learn Geant4), it may
be possible to train a generator directly on data to replace or augment
Full Simulation. This would require high fidelity testbeam data for a
variety of particles and energies. While this is beyond the scope of
this CWP, we suggest that this be pursued in parallel to improving the
speed and accuracy of fast simulators.

@@

In this note we have attempted to describe the main challenges and
opportunities for Full and Fast Simulation applications of HEP
experiments in view of planned future experimental programs in the next
decade; these include the HL-LHC, neutrino and muon experiments, and
studies towards future colliders such as FCC and CLIC. The combination
of additional physics models and higher precision in the simulation with
the need of larger Monte Carlo samples, which scale with the number of
real events recorded in future experiments, place an additional burden
on the computing resources that will be needed to generate them.

A multi-prong approach is being followed as a strategy for responding to
these challenges. Firstly Geant4 will continue to deliver new or refined
functionalities both in physics coverage and accuracy, whilst
introducing software performance improvements whenever possible. At the
same time, an R\&D programme is investigating new approaches that aim to
benefit from modern computing architectures. Its main feature is
track-level parallelisation, bundling particles with similar properties
from different events to process them in a single thread. This approach
combined with SIMD vectorisation coding techniques and use of data
locality is expected to yield large speed-ups, which are being measured
in a realistic prototype under development. In addition, the work on
Fast Simulation is accelerating with a view to producing a flexible
framework that permits Full and Fast simulation to be combined for
different particles in the same event. The overriding requirement is to
ensure the support of experiments through continuous maintenance,
support and improvement of the Geant4 simulation toolkit with minimal
API changes visible to these experiments at least until production
versions of potentially alternative engines, such as those resulting
from ongoing R\&D work, become available, integrated, and validated by
experiments. The agreed ongoing strategy to meet this goal is to ensure
that new developments resulting from the GeantV R\&D programme are
deployed in a timely way in Geant4.

>>>>

\abstract{ A working group on detector simulation was formed as part of the high-energy physics (HEP) 
Software Foundation's initiative to prepare a Community White Paper that describes 
the main software challenges and opportunities to be faced in the HEP field over the next 
decade. The working group met over a period of several months in order to review the 
current status of the Full and Fast simulation applications of HEP experiments 
and the improvements that will need to be made in order to meet the goals of future HEP 
experimental programmes.  The scope of the topics covered includes the main components 
of a HEP simulation application, such as MC truth handling, geometry modeling, particle 
propagation in materials and fields, physics modeling of the interactions of particles 
with matter, the treatment of pileup and other backgrounds, as well as signal processing 
and digitisation. The resulting work programme described in this document focuses on the need to 
improve both the software and the physics of detector simulation. The goals are to increase the 
accuracy of the physics models and expand their applicability to future physics programmes, 
while achieving large factors in computing performance gains consistent with projections on 
available computing resources.}

@@

For all its success so far, the challenges faced by the HEP field in the
simulation domain are daunting. During the first two runs, the LHC
experiments produced, reconstructed, stored, transferred, and analyzed
tens of billions of simulated events. This effort required more than
half of the total computing resources allocated to the experiments. As
part of the high-luminosity LHC physics program (HL-LHC) and through the
end of the 2030's, the upgraded experiments expect to collect 150 times
more data than in Run 1. Demand for larger simulation samples to satisfy
analysis needs will grow accordingly. In addition, simulation tools have
to serve diverse communities, including accelerator-based particle
physics research utilizing proton-proton colliders, neutrino and muon
experiments, as well as the cosmic frontier. The complex detectors of
the future, with different module- or cell-level shapes, finer
segmentation, and novel materials and detection techniques, require
additional features in geometry tools and bring new demands on physics
coverage and accuracy within the constraints of the available computing
budget. More extensive use of Fast Simulation is a potential solution,
under the assumption that it is possible to improve time performance
without an unacceptable loss of physics accuracy.

@@

Muon experiments such as Muon g-2 and Mu2e also face large simulation
challenges, being Geant4 the primary simulation toolkit for
all stages of these experiments. Because they are searching for extremely rare effects, they
must grapple with very low signal to background ratios, and the modeling
of low cross-section background processes. Often the physics of these
processes is not well understood, introducing large systematic
uncertainties. Additionally, the size of the computational problem is a
serious challenge, as large simulation runs are required to adequately
sample all relevant areas of experimental phase space, even when
techniques to minimise the required computations are used. Another
aspect related to the background events is the need to simulate the
effects of low energy neutrons, which requires large computational
resources. For example, Mu2e used a total of approximately 60 million
CPU hours, or 83 thousand core months, in a 2016 simulation campaign to
produce the Full Simulation sample needed to complete a technical design
report delivered in 2016.

Although in comparison the current computing needs of neutrino and muon experiments
are small compared to the LHC experiments, i.e. the CMS experiment
spent 860 thousand core months on simulation in the May 2015-May 2016
period, they are expected to grow substantially in the next decade.
It is therefore imperative that the
intensity frontier experiments make use of modern computing
architectures and algorithms. This transition will be complicated by the
fact that the intensity frontier is fractured into many small
experiments, some of which have as few as 50 collaborators, limiting the
manpower available for updating detector simulations to use new tools.
Efforts such as LArSoft{[}7{]} have recognised this challenge and aim to
pool resources of many experiments for the benefit of all.

@@

The Geant4 Collaboration has invested significant effort into improving
the toolkit computing performance in the past decade, as code was
reviewed and optimised. In 2013, the release of event-level
multithreading capabilities in Geant4 brought significant memory
savings, as illustrated in Figure 10, which shows the CPU time (top) and
memory (bottom) consumption of a simplified CMS standalone simulation
application, based on Geant4 version 10.3.p01, runs for 50 GeV pions.
Geant4 achieves almost perfect strong scaling (top plot) with the number
of available cores (57) and still gains about +30\% in hyperthreading
regime {[}29,30{]}. The large memory reduction obtained with
multi-threading is clearly visible (bottom). While a multi-process
approach can reach the O(10GB) memory consumption with a relatively
small number of threads (Geant4 version 9.6), it is only with a well
designed multi-threading approach, which shares the read-only data among
threads, that it is possible to run with the maximum number of threads
remaining in a memory budget of approximately 1 GB. Recent versions of
Geant4 have put emphasis on further reducing memory consumption to the
current level of about 10 MB needed for each additional thread.

@@

Tables 4 and 5 shows the relative CPU performance of a number of
configurations of the ATLAS simulation application. The largest gain
comes from introducing shower libraries in the forward calorimeter,
which comes at a cost of 430MB of VMEM per job. No other option applied
here has an appreciable effect on memory requirements of the
application. The neutron time cut shows only a small CPU improvement,
but was shown to reduce the size of the output file by almost a factor
of two in previous tests. This file size issue can also be mitigated by
various approaches to hit storage, however. Removing range cuts
decreases the simulation time per event, because the thin volumes of the
sampling calorimeter demand a tighter range cut, including 30~$\mu$m in the
forward calorimeter and 300~$\mu$m in the EM calorimeters, to provide an
accurate physics description. This penalty is not present in the CMS
simulation, since the CMS EM calorimeter has much larger active volumes.
Still, the effect of range cuts on the two applications is similar.

@@

There are many techniques to make simulation faster. Some of them are
applied to Geant4-based Full Simulation applications, such as for
example Russian Roulette and shower libraries, to speedup simulation in
a given sub-detector. These applications are still
labeled Full Simulation and Geant4-based. Some experiments have
developed alternative fast simulation frameworks, based on fast track
propagation through a simplified geometry and parameterised showers,
which are less accurate and are typically used in detector upgrade
studies and to produce ``signal'' samples that require scans of a large
region of theory parameter space. Generally, the faster the simulation,
the more imprecise the physics is. Examples of fast simulation
techniques, used to speedup Geant4-based applications or in fast
simulation frameworks, are:

@@

While the focus in this section and in the literature so far has been on
using machine learning to do fast simulation (e.g. learn Geant4), it may
be possible to train a generator directly on data to replace or augment
Full Simulation. This would require high fidelity test beam data for a
variety of particles and energies. While this is beyond the scope of
this CWP, we suggest that this be pursued in parallel to improving the
speed and accuracy of fast simulators.

@@

In this note we have attempted to describe the main challenges and
opportunities for Full and Fast Simulation applications of HEP
experiments in view of planned future experimental programs in the next
decade; these include the HL-LHC, neutrino and muon experiments, and
studies towards future colliders such as FCC and CLIC. The combination
of additional physics models and higher accuracy in the simulation with
the need of larger Monte Carlo samples, which scale with the number of
real events recorded in future experiments, place an additional burden
on the computing resources that will be needed to generate them.

A multi-prong approach is being followed as a strategy for responding to
these challenges. Firstly Geant4 will continue to deliver new or refined
functionalities both in physics coverage and accuracy, whilst
introducing software performance improvements whenever possible. At the
same time, an R\&D programme is investigating new approaches that aim to
benefit from modern computing architectures. For example, 
the main feature of the GeantV transport engine is
track-level parallelisation, bundling particles with similar properties
from different events to process them in a single thread. This approach
combined with SIMD vectorisation coding techniques and use of data
locality is expected to yield large speed-ups, which are being measured
in a realistic prototype under development. In addition, the work on
Fast Simulation is accelerating with a view to producing a flexible
framework that permits Full and Fast simulation to be combined for
different particles in the same event. The overriding requirement is to
ensure the support of experiments through continuous maintenance 
and improvement of the Geant4 simulation toolkit with minimal
API changes visible to these experiments at least until production
versions of potentially alternative engines, such as those resulting
from ongoing R\&D work, become available, integrated, and validated by
experiments. The agreed ongoing strategy to meet this goal is to ensure
that new developments resulting from the GeantV R\&D programme,
including the novel transport engine, are
deployed in a timely way in Geant4.
